Symmetric Circuits for Rank Gates

### 2018-02-11 sections/translating-formulas-to-cicuits.tex
Feb 10
feb 11

@@

@@

  We have from Anderson and Dawar~\cite{AndersonD17} that for every formula
  $\phi(\vec{x}) \in \FPC[\tau]$ there is a $\PT$-uniform family of symmetric
  circuits $(C^{\phi}_n)_{n \in \nats}$ with unique labels and symmetric gates
  that translates $\phi(\vec{x})$.

@@

  We thus derive the required $\PT$-uniform family $(\theta_n(\vec{x}))_{n \in
    \nats}$ by applying the first translation and then the second.

  Fix $n \in \nats$. We recursively define a function $T_n$ that maps $\psi
  (\vec{y})$, a sub-formula of $\theta_n(\vec{x})$, to a circuit $C^{\psi}_n$,
  where $C^{\psi}_n$ translates $\psi(\vec{y})$ at $n$. We then have that for
  all $n \in \nats$, $T_n(\theta_n(\vec{x}))$ translates $\theta_n$ (and hence
  $\theta$) at $n$.

@@

  Suppose $\psi(\vec{z}, \vec{w}, \vec{y})$ is a sub-formula of
  $\theta_n(\vec{x})$ such that $\psi (\vec{y}) = \rank^{\leq r}_p \vec{z},
  \vec{w} .\phi(\vec{z}, \vec{w}, \vec{y})$, for some $p, r \in \nats$ and $p$
  prime. Let $k:= \vert \vec{y} \vert$, $k_1 : = \vert \vec{z} \vert$ and $k_2
  := \vert \vec{w} \vert$. Let $R_\phi$ be a $(k_1 + k_2 + k)$-ary relation
  symbol not in $\tau$. Let $\psi'(\vec{y}) := \rank^{\leq r}_p\vec{z}, \vec{w}.
  R_{\phi} (\vec{z}, \vec{w}, \vec{y})$. Let $G_{\text{out}} := \{g_{\vec{c}} :
  \vec{c} \in [n]^k\}$ and $G_{\text{in}} : = \{ g_{\vec{a}, \vec{b}, \vec{c}} :
  \vec{a} \in [n]^{k_1}, \vec{b} \in [n]^{k_2}, \vec{c} \in [n]^{k}\}$. Let
  $C^{\psi'}_n:= \langle G^{\psi'}, \Omega^{psi'}, \Sigma^{\psi'},
  \Lambda^{\psi'}, L^{\psi'}\rangle$ be a $(\mathbb{B}_{\rank}, \{ R_{\phi}
  \})$-circuit such that
  \begin{itemizens}
  \item $G^{\psi'} = G_{\text{out}} \cup G_{\text{in}}$,
  \item $\Omega^{\psi'}$ be an injection from $[n]^k$ to $G^{\psi'}$ defined by
    $\Omega^{\psi'}(\vec{c}) = g_{\vec{c}}$ for all $\vec{c} \in [n]^k$,
  \item $\Sigma^{\psi'} (g) = \rank^r_p$ if $g \in G_{\text{out}}$ and
    $\Sigma^{\psi'}(g) = R_\phi$ if $g \in G_{\text{in}}$,
  \item for each $(\vec{a}, \vec{b}, \vec{c}) \in [n]^{k_1 +k_2 + k}$,
    $\Lambda^{\psi'} (\vec{a}, \vec{b}, \vec{c}) = g_{\vec{a}, \vec{b},
      \vec{c}}$, and
  \item for each gate $g_{\vec{c}} \in G_{\text{out}}$, $L(g_{\vec{c}}):
    [n]^{k_1} \times [n]^{k_2} \rightarrow G^{\psi'}$ is defined by
    $L(g_{\vec{c}})(\vec{a}, \vec{b}) := g_{\vec{a}, \vec{b}, \vec{c}}$ for each
    $\vec{a} \in [n]^{k_1}$ and $\vec{b} \in [n]^{k_2}$.
  \end{itemizens}

>>>>

\begin{definition}
  The function $\op{tag-inputs}$ maps a circuit $C := \langle G, \Omega, \Sigma,
  \Lambda, L \rangle$, a relation symbol $R \in \tau$, and $H \subseteq G$
\end{definition}

@@

We say that a circuit $C$ \emph{translates} a formula $\psi(\vec{x})$ for $n \in
\nats$ if $C$ is a transparent symmetric rank-circuit of order $n$, $C$ is
defined over the same vocabulary as $\psi (\vec{x})$, and $C$ computes the query
defined by $\psi(\vec{x})$ on structures of size $n$. We say a family of
circuits $(C_n)_{n \in \nats}$ \emph{translates} a formula $\psi (\vec{x})$ if
$(C_n)_{n \in \nats}$ is $\PT$-uniform and for all $n \in \nats$, $C_n$
translates $\psi(\vec{x})$ for $n$.

\begin{definition}
	Let $\alpha : X \rightarrow Y$ and $\beta : W \rightarrow Z$. We say that $f$
  is \emph{compatible} with $g$, and we write $f \sim g$, if for all $a \in X
  \cap W$, $f(a) = g(a)$ and for all $a \in X \setminus W$ and $b \in W
  \setminus X$, $f(a) \neq g(b)$.
\end{definition}

We now prove that for every $\FOrk$-formula $\theta(\vec{x})$ and $n \in \nats$
there exists a circuit $C$ that translates $\theta(\vec{x})$ for $n$. The gates
of $C$ are of the form $g_{\psi, \alpha}$, where $\psi(\vec{y})$, a sub-formula
of $\theta(\vec{x})$, and $\alpha$ is an assignment to the free variables of
$\psi$. We then wire up the gates such that $g_{\psi, \alpha}$ is associated
with an appropriate function from the basis, and the children of $g_{\psi,
  \alpha} $ are those gates $g_{\phi, \beta}$, where $\phi$ is an immediate
sub-formula of $\psi$ and $\beta$ is an assignment to the free variables in
$\phi$ that agrees with the assignment $\alpha$. The resultant circuit is
symmetric and has the property that $g_{\psi, \alpha}$ evaluates to one for a
given input structure if, and only if, $\psi$ holds for that structure given the
assignment $\alpha$. However, as stated this construction does not result in a
transparent circuit. In order to ensure that $C$ is also transparent we first
alter $\theta$ by appending to each application of the rank operator a gadget
that ensures that no two children of a rank gate are syntactically-equivalent.
We now formally define this construction and, moreover, provide a bound on the
running time of the algorithm constructing this circuit.

\begin{lem}
  There is an algorithm that takes in a $\FOrk$-formula $\theta(\vec{x})$ and $n
  \in \nats$ and outputs a rank-circuit $C$ that translates $\theta(\vec{x})$
  for $n$. Moreover, there is a polynomial in $p$ such that for an input
  $(\theta(\vec{x}), n)$ the algorithm terminates in at most
  $p(n^{\width(\theta)}\vert \cl{\theta} \vert)$ many steps.
\end{lem}
\begin{proof}
  Let $\tau$ be the vocabulary of $\theta(\vec{x})$. Let $k := \width(\theta)$.
  Suppose $\tau$ is empty. In this case we can evaluate $\theta$ for each
  assignment $\vec{a}$ in a number of steps polynomial in $n^k \vert \cl{\theta}
  \vert$. Let $f : [n]^k \rightarrow \{0,1\}$ be defined such that $f (\vec{a})
  = 1$ if, and only if, $[n] \models \theta[\vec{a}]$. Let $C = \langle G,
  \Sigma, \Omega, \Lambda, L \rangle$, where
  \begin{itemizens}
  \item $G = \{g_0, g_1\} \cup \{g_{\vec{a}} : \vec{a} \in [n]^{k}\}$,
  \item $\Sigma (g_0) = 0$, $\Sigma (g_1) = 1$ and $\Sigma (g_{\vec{a}}) =
    \land[1]$,
  \item $\Omega : [n]^k \rightarrow G$ is defined by $\Omega(\vec{a}) =
    g_{\vec{a}}$ for all $\vec{a} \in [n]^k$,
  \item $\Lambda = \emptyset$,
  \item for $\vec{a} \in [n]^k$, $L(g_{\vec{a}}) : [1] \rightarrow G$ is defined
    by $L(g_{\vec{a}})(1) = g_{f(\vec{a})}$.
  \end{itemizens}

  Since $C$ has symmetric gates, $C$ is transparent. Notice that, since
  $\theta(\vec{x})$ defines a query on structures over the empty vocabulary, we
  have for all $\sigma \in \sym_n$, $f(\sigma \vec{a}) = f(\vec{a})$ for all
  $\vec{a} \in [n]^k$. For $\sigma \in \sym_n$, let $\pi_\sigma : G \rightarrow
  G$ be a bijection that fixes the constant gates and such that $\pi_{\sigma}
  (g_{\vec{a}}) = g_{\sigma \vec{a}}$ for all $\vec{a} \in [n]^k$. Then
  $\pi_\sigma \Omega (\vec{a}) = \pi_\sigma g_{\vec{a}} = g_{\sigma \vec{a}} =
  \Omega (\sigma \vec{a})$. It follows that $\pi_\sigma$ is an automorphism of
  the circuit extending $\sigma$. It follows that $C$ is symmetric, and so $C$
  translates $\theta(\vec{x})$ for $n$.

  Suppose $\tau$ is not empty. Let $R \in \tau$. We now construct a $C$ that
  translates $\theta(\vec{x})$ at $n$. We first pre-proccess the formula $\theta(\vec{x})$

  Let $\op{no-op}(y) := (R(y, y) \land (\neg R(y, y)))$. Let $\vec{y} := (y_1,
  \ldots, y_k)$ be a vector of variables and $\op{add-tag} (\vec{y}) :=
  \op{no-op}(y_1) \land (\op{no-op}(y_2) \land ( \op{no-op}(y_2) \land ( \ldots
  \land (\op{no-op}(y_k)) \ldots)))$. From $\theta (\vec{x})$ we construct a new
  formula $\theta' (\vec{x})$ by replacing each sub-formula $\phi(\vec{y})$ of
  the form $\rank^r_p \vec{w} \vec{z} . \psi$ by the sub-formula $\rank^r_p
  \vec{w}\vec{z} . ((\forall u . u = u) \land \phi) \land \op{add-tag}(\vec{w}
  \cup \vec{z})$.

  Fix $n \in \nats$. For each $\psi \in \cl{\theta'}$ let $G_\psi := \{g_{\psi,
    \alpha} : \alpha \in [n]^{\free{\psi}} \}$. Let $G_c := \{g_0, g_1\}$. Let
  $G := G_c \cup (\bigcup_{\psi \in \cl{\theta'}} G_{\psi})$. We define
  $\Sigma$, $\Lambda$ and $L$ as follows. Let $\Sigma (g_1) = 1$ and $\Sigma
  (g_0) = 0$. If $g_{\psi, \alpha} \in G \setminus G_c$.

  \begin{itemizens}
  \item If $\psi$ is of the form $T(\vec{y})$ for some $T \in \tau$, then
    $\Sigma (g_{\psi, \alpha}) = T$ and $\Lambda_T (g_{\psi, \alpha}) = \alpha
    \vec{y}$.
  \item If $\psi$ is of the form $(y_1 = y_2)$, then $\Sigma (g_{\psi, \alpha})
    = \land[1]$ and $L(g) : [1] \rightarrow G$ by $L(g)(1) = g_1$ if $\alpha
    (y_1) = \alpha (y_2)$ and $L(g)(1) = g_0$ otherwise.
  \item If $\psi$ is of the form $\rank^r_p \vec{w} \vec{z}. \phi$, let $a =
    n^{\vert \vec{w}\vert}$ and $b = n^{\vert \vec{z}\vert}$, and $\Sigma
    (g_{\psi, \alpha}) = \rank[a, b]$. Let $L(g) : [a] \times [b] \rightarrow G$
    be defined by $L(g)(i,j) = g_{\phi, \beta}$, where $\beta : \vec{w} \cup
    \vec{z} \cup \vec{y} \rightarrow [n]$ is such that $\alpha \subseteq \beta$
    and such $\beta \vec{y}$ and $\beta \vec{z}$ are the $i$th and $j$th
    elements of $[n]^{\vert \vec{w} \vert}$ and $[n]^{\vert \vec{z} \vert}$ in
    the lexicographical ordering.
  \item If $\psi$ is of the form $Q z . \phi(\vec{y}, z)$, for $Q \in \{\forall,
    \exists\}$, then if $Q = \forall$ we have $\Sigma (g_{\psi, \alpha}) = \land
    [n]$, otherwise $\Sigma (g_{\psi, \alpha}) = \lor [n]$. Let $L(g) : [n]
    \rightarrow G$ be defined by $L(g)(i) = g_{\phi, \beta_i}$, where $\beta_i
    \in [n]^{\vec{y} \cup \{z\}}$ and $\alpha \subseteq \beta_i$ and $\beta_i
    (z) = i$.
  \item If $\psi$ has free variables $\vec{y}$ and is of the form $\phi_1
    (\vec{y}_1) \land \phi_2(\vec{y}_2)$, where $\vec{y_1}, \vec{y_2} \subseteq
    \vec{y}$, then $\Sigma(g_{\psi, \alpha}) = \land[2]$ and $L(g) : [2]
    \rightarrow G$ is defined by $L(g)(i) = g_{\phi_i, \beta_i}$ where $\beta_i
    \in [n]^{\vec{y}_i}$ and $\beta_i \subseteq \alpha$ for $i \in [2]$. The
    same approach is used in the disjunctive case.
  \item if $\psi$ is of the form $\neg \phi$ then $\Sigma (g_{\psi, \alpha}) =
    \neg$ and $L(g): [1] \rightarrow G$, where $L(g)(1) = g_{\phi, \alpha}$.
  \end{itemizens}

  We let $\Omega : [n]^k \rightarrow G$ be defined by $\Omega (\vec{a}) =
  g_{\theta', \alpha}$, where $\alpha \vec{x} = \vec{a}$, for all $\vec{a} \in
  [n]^k$. Let $C := \langle G, \Omega, \Sigma, \Lambda, L \rangle$. We notice
  that $\vert C \vert \leq n^{\width(\theta')} \vert \cl{\theta} \vert$. It can
  be shown that $C$ can be constructed in time bounded by a polynomial in the
  size of $C$.

  We now show that $C$ translates $\theta(\vec{x})$ for $n$. It is obvious that
  $C$ has order $n$. Let $\sigma \in \sym_n$. Let $\pi_\sigma : G \rightarrow G$
  be a defined such that $\pi (g) = g$ if $g \in G_c$ and $\pi_\sigma(g_{\psi,
    \alpha}) = g_{\psi, \sigma \alpha}$ for all $\psi \in \cl{\theta}$ and
  $\alpha \in n^{\free(\psi)}$. It is easy to see that $\pi_\sigma$ is an
  automorphism of $C$ extending $\sigma$. It follows that $C$ is symmetric. Let
  $g := g_{\psi, \alpha} \in G$ be a rank gate and let $h, h' \in H_g$. Suppose
  $\psi (\vec{y}) := \rank^r_p \vec{w} \vec{z}. \phi$ and $h := g_{\phi, \beta}$
  and $h' := g_{\phi, \beta'}$ are syntactically-equivalent. From the
  construction of $\theta'$ we have that $\phi$ must be of a particular form.
  Suppose $\phi = \eta \land \op{add-tag}(\vec{w} \cup \vec{z})$. Let $\gamma$
  and $\gamma'$ be the restrictions of $\beta$ and $\beta'$ to $\vec{w} \cup
  \vec{z}$, respectively, and let $\zeta(\vec{w}\cup\vec{z}) :=
  \op{add-tag}(\vec{w} \cup \vec{z})$. The $\op{add-tag} (\vec{w} \cup \vec{z})$

  The structure of this formula gives us that $h$ is syntactically-equivalent to
  $h'$ if, and only if, $g_{\zeta , \beta'}$

  We also note that $C$ has order $n$.

\end{proof}

@@

  % We have from Anderson and Dawar~\cite{AndersonD17} that for every formula
  % $\phi(\vec{x}) \in \FPC[\tau]$ there is a $\PT$-uniform family of symmetric
  % circuits $(C^{\phi}_n)_{n \in \nats}$ with unique labels and symmetric gates
  % that translates $\phi(\vec{x})$.

@@

  We thus derive the required $\PT$-uniform family $(\theta_n(\vec{x}))_{n \in
    \nats}$ by applying the first translation and then the second. Let $x$ be a
  variable and $R \in \tau$. Let $\op{no-op}(x) := R(x, x) \land (\neg R(x,
  x))$. Let $\vec{x} := (x_1, \ldots, x_k)$ and $\op{add-tag} (\vec{x}) :=
  \op{no-op}(x_1) \land (\op{no-op}(x_2) \land ( \ldots \land (\op{no-op}(x_k))
  \ldots)$. For each $\theta_n (\vec{x})$ we construct a new formula $\theta_n'
  (\vec{x})$ by replacing each sub-formula $\phi(\vec{y})$ of
  $\theta_n(\vec{x})$ of the form $\rank^r_p \vec{w} \vec{z} . \psi$ by
  $\rank^r_p \vec{w}\vec{z} . (\psi \land \op{add-tag}(\vec{w} \cup \vec{z}))$.
  It is easy to see that we can construct $\theta_n'$ from $\theta_n$ in time
  polynomial in $\theta_n$. Moreover, $\theta_n'$ and $\theta_n$ are equivalent
  and have the same width.

  Fix $n \in \nats$. We now construct a circuit $C_n$ that translates
  $\theta_n'$ (and hence $\theta_n$ and $\theta$) at $n$. For each $\psi \in
  \cl{\theta_n}$ let $G_\psi := \{g_{\psi, \alpha} : \alpha \in
  [n]^{\free{\psi}} \}$. Let $G := \bigcup_{\psi \in \cl{\theta_n}} G_{\psi}$.
  Let $g_{\psi, \alpha} \in G$. We define $\Sigma$, $\Lambda$ and $L$ as
  follows.
  \begin{itemizens}
  \item If $\psi (\vec{y}) = R(\vec{y})$, for some $R \in \tau$, then $\Sigma
    (g_{\psi, \alpha}) = R$ and $\Lambda_R (g_{\psi, \alpha}) = \alpha
    \vec{y})$.
  \item If $\psi (y_1, y_2) = (y_1 = y_2)$, then if $\alpha (y_1) = \alpha
    (y_2)$, $\Sigma (g_{\psi, \alpha}) = 1$, else $\Sigma (g_{\psi, \alpha}) =
    0$.
  \item if $\psi (\vec{y}) = \rank^r_p \vec{w} \vec{z}. \phi$, then $\Sigma ()$
  \end{itemizens}

@@

  Suppose $\psi(\vec{z}, \vec{w}, \vec{y})$ is a sub-formula of
  $\theta_n(\vec{x})$ such that $\psi (\vec{y}) = \rank^{\leq r}_p \vec{z},
  \vec{w} .\phi(\vec{z}, \vec{w}, \vec{y})$, for some $p, r \in \nats$ and $p$
  prime. Let $k:= \vert \vec{y} \vert$, $k_1 : = \vert \vec{z} \vert$ and $k_2
  := \vert \vec{w} \vert$. Let $R_\phi$ be a $(k_1 + k_2 + k)$-ary relation
  symbol not in $\tau$. Let $\psi'(\vec{y}) := \rank^{\leq r}_p\vec{z}, \vec{w}.
  R_{\phi} (\vec{z}, \vec{w}, \vec{y})$. Let $G_{\text{out}} := \{g_{\vec{c}} :
  \vec{c} \in [n]^k\}$ and $G_{\text{in}} : = \{ g_{\vec{a}, \vec{b}, \vec{c}} :
  \vec{a} \in [n]^{k_1}, \vec{b} \in [n]^{k_2}, \vec{c} \in [n]^{k}\}$. Let
  $C^{\psi'}_n:= \langle G^{\psi'}, \Omega^{\psi'}, \Sigma^{\psi'},
  \Lambda^{\psi'}, L^{\psi'}\rangle$ be a $(\mathbb{B}_{\rank}, \{ R_{\phi}
  \})$-circuit such that
  \begin{itemizens}
  \item $G^{\psi'} = G_{\text{out}} \cup G_{\text{in}}$,
  \item $\Omega^{\psi'}$ be an injection from $[n]^k$ to $G^{\psi'}$ defined by
    $\Omega^{\psi'}(\vec{c}) = g_{\vec{c}}$ for all $\vec{c} \in [n]^k$,
  \item $\Sigma^{\psi'} (g) = \rank^r_p$ if $g \in G_{\text{out}}$ and
    $\Sigma^{\psi'}(g) = R_\phi$ if $g \in G_{\text{in}}$,
  \item for each $(\vec{a}, \vec{b}, \vec{c}) \in [n]^{k_1 +k_2 + k}$,
    $\Lambda^{\psi'} (\vec{a}, \vec{b}, \vec{c}) = g_{\vec{a}, \vec{b},
      \vec{c}}$, and
  \item for each gate $g_{\vec{c}} \in G_{\text{out}}$, $L(g_{\vec{c}}):
    [n]^{k_1} \times [n]^{k_2} \rightarrow G^{\psi'}$ is defined by
    $L(g_{\vec{c}})(\vec{a}, \vec{b}) := g_{\vec{a}, \vec{b}, \vec{c}}$ for each
    $\vec{a} \in [n]^{k_1}$ and $\vec{b} \in [n]^{k_2}$.
  \end{itemizens}
